---
title: "Final Report"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 4
    theme: lumen
---

```{r, include=FALSE}
library(tidyverse)
library(tidytext)  
library(httr)
library(rvest)
library(lubridate)
library(plotly)
library(patchwork)
library(lmtest)
library(dplyr)
library(tidycensus)
require(data.table)
library(gridExtra)
library(ggplot2)
library(data.table)
library(tidycensus)
library(tidyverse)
library(ggridges)
library(patchwork)
library(viridis)
library(modelr)
library(mgcv)
library(readxl)
library(plotly)
library(kableExtra)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


## Motivation

In 2020, the world entered into a pandemic and now two years later as we recover and try to move on with life, things are different. The pandemic brought with it a new lens that showed disparities based on SES, geographical location, etc. that always existed but were never taken seriously. Research has shown that people of color and those from a low SES are at higher risk to contracting COVID-19; as such, public health professionals should be aware of such gaps and make it an effort to especially focus on these groups at higher risk (Khanijahani et al., 2021).

## Analysis and Relevance:

Our project looks at the change in COVID-19 trends in New York State for each county, specifically how COVID-19 is spatially autocorrelated across the different counties and the extent auto-correlation within New York City, the most populated county. We also looked at whether health behaviors and socioeconomic (SES) factors are associated with this change in trends by county. These analyses can be helpful to public health officials in the following ways:
  
First, understanding the spatial patterns of COVID-19 trends can help public health officials to identify areas that are at higher risk for the spread of the virus. This information can be used to target interventions and resources to these areas in order to reduce the spread of COVID-19 and protect the health of the population.

Second, examining the relationship between health behaviors and SES factors and the change in COVID-19 trend can help public health officials to identify potential risk factors for the spread of the virus. This information can be used to develop targeted interventions and policies that aim to reduce the risk of COVID-19 transmission by addressing these factors.

Overall, studying the change in COVID-19 trend in New York State for each county, and examining spatial autocorrelation and the relationship with health behaviors and SES factors, can provide valuable insights. These discoveries can help public health officials to better understand the spread of COVID-19 and take effective action to reduce the risk of transmission in communities at higher risk and address presisting health disparities.

## Data Processing and Cleaning

### Data Sources
Our data were downloaded from four different sources:

1. **New York State Demographics**: US Census Bureau's demographics data were imported using `tidycensus`.

2. **New York State, County Land Area** [(dataset)](https://www.health.ny.gov/statistics/vital_statistics/2010/table02.htm)
A .csv file was created based on the information provided in the website.  
Note: unable to be loaded using `read_html` as this is an xml document.

3. **New York State Health Ranking** [(dataset)](https://www.countyhealthrankings.org/explore-health-rankings/new-york/data-and-resources):
Four data sets (2019, 2020, 2021 and 2022 New York State Health Ranking Data) were downloaded as .xls/xlsx. These data sets are provided by the University of Wisconsin Population Health Institute, which performs extensive analyses to calculate each New York county's score for health factors (health behaviors, clinical care, SES factors, and physical environment) and health outcome (quality of life, and length of life). 

4. **New York State COVID-19** [(dataset)](https://data.ny.gov/browse?tags=covid-19):
Three data sets (COVID testing, COVID vaccination, COVID fatalities from 2020 to 11/2022) were downloaded as .csv.

### Data Cleaning
1. **New York State Demographics**  
We imported US Census Bureau's 2020 data because 2021 data are not available until 12/8/2022. As demographics data do not change significantly each year, 2020 data were used for analysis. Then we decided to include the following variables, as we would like to adjust them in our models:
* total population in each county (tpop)
* median age in each county (medage)
* racial diversity index (race_div): calculated Simpson's Diversity Index in lines 65 to 69 (higher index indicates more racial diversity)

```{r census_data, results = FALSE, message = FALSE}
census_api_key("aa512886c5449a582d837da8d3a07af66a043fe5")

census_data <- load_variables(2020, "acs5", cache=T)
fwrite(census_data, "census_variables.csv")

vars <- c(tpop = 'P001001',
          medage = 'P013001',
          wpop = 'P003002',
          bpop = 'P003003',
          apop = 'P003005',
          hpop = 'P004003')

nys_df <- get_decennial(state = "ny", 
                       geography = "county",
                       variables = vars,
                       geometry = T,
                       output = "wide")

NY_df <- get_decennial(state = "ny", 
                       geography = "county",
                       variables = vars,
                       geometry = T,
                       output = "wide")

NY_df = NY_df %>%
  mutate(
    county = gsub(" County, New York","",NAME))

nys_df = nys_df %>%
  mutate(
    county = gsub(" County, New York","",NAME))

nys_df$race_div <- 1 - (((nys_df$wpop*(nys_df$wpop-1))+
                          (nys_df$bpop*(nys_df$bpop-1))+
                          (nys_df$hpop*(nys_df$hpop-1))+
                          (nys_df$apop*(nys_df$apop-1)))/
                         (nys_df$tpop*(nys_df$tpop-1)))

nys_df = nys_df %>% 
  select(tpop, medage, county, race_div)
```

2.  **New York State, County Land Area**
After loading the dataset, which only contains two variables--county and value (provides area of each county as square mile; this variable was renamed as "area")--the dataset was merged with nys_df to calculate population density for each county. The population density variable was created so we can adjust for it in our regression models, especially ones that involve number of COVID-19 cases and fatalities.

```{r, results = FALSE}
area = read_excel('data/pop_density.xlsx') %>% 
  janitor::clean_names() %>% 
  rename(area = value)

nys_df = left_join(nys_df, area, by = 'county') %>% 
  mutate(pop_d = tpop/area)
```

3. **New York State Health Ranking**  
Same steps were taken for each year except 2019; for this year, the sub-variables from the "Ranked Measure Data" sheet were labeled slightly different compared to the follow years so for consistency, each of the successive years, the variables needed to be renamed to match the original names from the 2019 dataset: 
As the original excel files contain multiple sheets. We only imported the sheets that contain variables of interest.  
 <br>Each county's health ranking was imported from the sheet named "Outcomes & Factors SubRankings." From the sheet, only the following variables were selected for future analysis: each county's rank for health outcomes (quality of life, renamed to qol_r; length of life, renamed to longevity_r), health factors (health behaviors, renamed to health_beh_r; clinical_care, renamed to clinical_care_r; socioeconomic factors, renamed to ses_r; physical environment, renamed to env_r).
 <br>Additionally, further information on each county's socioeconomic factors and race/ethnicity were imported from the sheet named "Additional Measure Data." The following variables were selected: each county's segregation index (white vs. nonwhite), renamed to segregation_score; median household income, renamed household_income; % Hispanic, p_hispanic; % black, renamed p_black). As we are interested in observing racial and socioeconomic disparities in COVID, we then created the following variables: % minority (hispanic + black), named p_minority; binary variable for low vs. moderate/high level of segregation. The cutoff value for level of segregation was set as 30. [(source)](https://www1.udel.edu/uapp800/Lecture%20Material/Index%20of%20Dissimilarity%20Example.htm)
 <br>Lastly, we imported data from the sheet labeled "Ranked Measure Data" to import the sub-variables that make up the Health Behaviors Ranking. We imported the variables county, percent of smokers, percent of adults with obesity, food environment index (0-10 with 10 as the best), percent of adults who are physically inactive, percent of adults with access to location with opportunities to exercise, percent of adults who report excessive drinking, percent of driving related deaths with alcohol involved, the number of chlamydia cases, and the teen birth rate per every 1000 females aged 15-19.

```{r, results = FALSE, message = FALSE} 
# 2019 HR Data

hr2019 = read_excel('data/ny_hr19.xls', 
                    sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2019 = read_excel('data/ny_hr19.xls', 
                      sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'segregation_index_2',
         'household_income',
         'percent_hispanic',
         'percent_african_american') %>% 
  rename(median_income = household_income,
         segregation_score = segregation_index_2,
         p_hispanic = percent_hispanic,
         p_black = percent_african_american) %>%
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var19 =
  read_excel('data/ny_hr19.xls', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_obese, food_environment_index, percent_physically_inactive, percent_with_access, percent_excessive_drinking, percent_alcohol_impaired, number_chlamydia_cases, teen_birth_rate, percent_unemployed)
  

list_2019 =
  list(hr2019, demo2019, sub_var19)

total2019 = 
  list_2019 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2019)

# 2020 HR Data
hr2020 = read_excel('data/ny_hr20.xlsx', 
                    sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2020 = read_excel('data/ny_hr20.xlsx', 
                      sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'median_household_income', 
         'segregation_index_2',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index_2,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var20 =
  read_excel('data/ny_hr20.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)
  

list_2020 =
  list(hr2020, demo2020, sub_var20)

total2020 = 
  list_2020 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2020)

# 2021 HR Data
hr2021 = read_excel('data/ny_hr21.xlsx', 
                    sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2021 = read_excel('data/ny_hr21.xlsx', 
                      sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'median_household_income', 
         'segregation_index_2',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index_2,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var21 =
  read_excel('data/ny_hr21.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)

list_2021 =
  list(hr2021, demo2021, sub_var21)

total2021 = 
  list_2021 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2021)

# 2022 HR Data
hr2022 = read_excel('data/ny_hr22.xlsx', 
                    sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2022 = read_excel('data/ny_hr22.xlsx', 
                      sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
    select('county', 
         'median_household_income', 
         'segregation_index',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_hispanic + p_black)

sub_var22 =
  read_excel('data/ny_hr22.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)

list_2022 =
  list(hr2022, demo2022, sub_var22)

total2022 = 
  list_2022 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2022)

# Merge
hr_all = rbind(total2019, total2020, total2021, total2022)
```

3. **New York State COVID-19**  
From the three data sets we extracted the following variables of interest: number of new COVID cases per day, cumulative number of deaths reported per day, cumulative # of individuals who received the first dose of the vaccine. For the data set with cumulative number of deaths, the county "Manhattan" was replaced with "New York" for consistency. Next, the three data sets were merged by "date" and "county." The merged data set was then grouped according to "county" and "year" and the following variables were created: total number of new positive cases per year, called n_cases; total number of deaths per year, called n_deaths; and total number of individuals who received the first dose of the vaccine per year, called vax_dose1 (this data is not available for year 2020).

```{r, results = FALSE, message = FALSE, warnings = FALSE}
ny_test = read_csv('data/ny_covidtest.csv') %>% 
janitor::clean_names() %>% 
filter(geography == 'COUNTY') %>% 
select(-geography) %>% 
rename(date = test_date) %>% 
mutate(date = lubridate::mdy(date))  

ny_death = read_csv("data/ny_coviddeaths.csv") %>% 
janitor::clean_names() %>% 
rename(date = report_date, n_fatality = place_of_fatality) %>%
select(-deaths_by_county_of_residence) %>% 
mutate(date = lubridate::mdy(date), 
       county = replace(county, county == 'Manhattan', 'New York'))

ny_vax = read_csv("data/ny_covidvax.csv") %>% 
janitor::clean_names() %>% 
rename(date = report_as_of) %>% 
select(-region, -series_complete) %>% 
mutate(date = lubridate::mdy(date))

ny_covid_v1 = left_join(ny_test, ny_death, by = c("date", "county"))

covid_df_v1 = left_join(ny_covid_v1, ny_vax, by = c("date", "county")) %>% 
  separate(date, c("year", "month", "day"), sep = "-") %>% 
  mutate(year = as.numeric(year), 
         month = as.numeric(month), 
         day = as.numeric(day))

covid_df = covid_df_v1 %>% 
  group_by(year, county) %>% 
  summarise(n_cases = sum(new_positives),
            vax_dose1 = max(first_dose),
            n_deaths = max(n_fatality, na.rm = TRUE))
```

4. **Dataset for Analysis**  
For the first section of regression tests, the demographics dataset, HR dataset, and COVID dataset were merged into one dataset called "comb." At the end, we added a new binary variable called vax_maj to see whether majority of the residents in the county had received the first dose of vaccine. The cutoff value used was 79% as it was reported that nearly ~79% of the US population had completed at least one dose of vaccine by the end of 2022. [(source)](https://usafacts.org/visualizations/covid-vaccine-tracker-states/state/new-york)  

```{r, results = FALSE, message = FALSE}
covid_hr = left_join(covid_df, hr_all, by = c("county", "year")) 
comb = left_join(covid_hr, nys_df, by = 'county') %>% 
  mutate(tpop_50 = tpop/2,
         vax_maj = case_when(vax_dose1 >= .83*(tpop) ~ 1,
                             vax_dose1 < .83*(tpop) ~ 0))
```

5. **Dataset for Additional Analysis **  
For the additional regression analysis section, in order to have a more focused dataset, a new dataset was created that is a combination of the Census demographics, NY COVID data, and the Sub-variables dataset from the "Ranked Measure Data" sheet for each of the four Health Ranking datasets. 

```{r, results = FALSE, message = FALSE}

sub_var_data =
  hr_all %>%
  select(-ends_with("_r"))

covid_hr = left_join(covid_df, sub_var_data, by = c("county", "year")) 

new_comb = left_join(covid_hr, nys_df, by = 'county')

```

5. **Bubble Plot**  
We were interested in seeing whether there is a correlation between SES factors ranking and health behaviors and how it varies by the total number of residents in each county.

```{r}
bubble_df = comb %>% 
  select(year, county, health_beh_r, ses_r, tpop) %>% 
  filter(year == '2021')
```

6. **Bar Graphs for EDA**  
County health factors ranking (HR) was categorized into 4 quartiles quartile 1 = ranks 1 to 16; quartile 2 = ranks 17 to 31; quartile 3 = ranks 32 to 46, quartile 4 = ranks 47 to 62).[(source)](https://www.countyhealthrankings.org/reports/state-reports/2022-new-york-state-report) To observe any patterns between HR and COVID.

```{r}
eda1 = comb %>% 
  pivot_longer(health_beh_r:env_r,
               names_to = "rank_var",
               values_to = "rank") %>% 
  filter(year == '2021') %>% 
  mutate(quartile = case_when(rank < 17 ~ 'q1',
                              rank < 32 ~ 'q2',
                              rank < 46 ~ 'q3',
                              rank < 63 ~ 'q4')) %>% 
  mutate(quartile = as.factor(quartile))
```


## Exploratory Data Analysis

In order to have a better understanding of the existing socioeconomic and health factor status of the counties we conducted some exploratory data analysis mainly through visualization. We first tried to see which counties have the highest socio-economic and heath factor scores

```{r}
library(plotly)
data = bubble_df %>%
  mutate(county = as.factor(county),
         health_beh_r = as.numeric(health_beh_r),
         ses_r = as.numeric(ses_r)) %>%
  janitor::clean_names()


total <- merge(bubble_df,NY_df,by="county")

fig <- plot_ly(total, x = ~health_beh_r, y = ~ses_r, text = ~county, type = 'scatter',
               mode = 'markers', size = ~tpop.y, color = ~county, colors = 'Paired',
               marker = list(opacity = 0.5, sizemode = 'diameter'))
fig <- fig %>% layout(title = 'health-socio',
                      xaxis = list(showgrid = FALSE),
                      yaxis = list(showgrid = FALSE),
                      showlegend = FALSE)

fig
write.csv(total,"./data/eda_bubble.csv")

```

From the plot we can see that New York, Queens and Kings have the highest socio economic score, but have the lowest health factors ranks. The size of the bubble signifies total population of the county. Nassau, Suffolk, Westcher have low health and socio-ecnomic scores. 

Below is a colorpleth map to good understanding on the above in a spatial context,: 

```{r}
#library(sf)
#total = st_as_sf(total)
p = list()
p[[1]] = total %>%
  ggplot() +
  geom_sf(aes(fill = medage, geometry = geometry)) + 
  scale_fill_viridis_c(option = "viridis") +
  ggtitle("Median Age Per County New York State")


p[[2]] = total %>%
  ggplot() +
  geom_sf(aes(fill = health_beh_r, geometry = geometry)) + 
  scale_fill_viridis_c(option = "viridis") +
  ggtitle("Heath Rank Per County New York State")

grid.arrange(p[[1]], p[[2]], nrow = 1)

```

On the left we can see a chloropleth mao of medage per county in new york on the right is a chloropleth map of health factor rank per county.

Before testing association between county health factors (HR) and COVID, we wanted to see rather there are any observable patterns between COVID cases & HR, COVID vaccination & HR, and COVID deaths & HR.
```{r, message = FALSE, warning = FALSE}
p1 = ggplot(data=eda1, aes(x=rank_var, y=n_deaths, fill=quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Reds") + 
  theme_minimal() +
  labs(title="Number of COVID Deaths and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of Deaths")

p2 = ggplot(data=eda1, aes(x=rank_var, y=n_cases, fill=quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Oranges") + 
  theme_minimal() +
  labs(title="Number of COVID Cases and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of Cases")

p3 = ggplot(data=eda1, aes(x=rank_var, y=vax_dose1, fill=quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Greens") + 
  theme_minimal() +
  labs(title="Covid Vaccination (Dose 1) and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of People Vaccinated")

grid.arrange(p1, p2, p3, ncol = 1)
```

## Statistical Analysis

Note: Though we initially planned to adjust for racial diversity in our model, the variable was removed because racial diversity index did not vary much among counties. Also, tpop (total population) was replaced with pop_d (population density) as it was shown that COVID-19 incidence is strongly correlated with population density (Iderus, 2022). For the models, only data from 2021 or 2022 were used as there is no information on vaccination for 2020.

<p style="color: red">***Question 1: Are county health factors associated with COVID-19 risk?***</p>
Health factors--health behaviors, clinical care, physical environment, and socioeconomic factors--are community conditions that can be modified and improved through public health interventions or health policies. Therefore, we were interested in seeing which factor(s) is significantly associated with the number of COVID-19 cases, deaths, and vaccination. To do so, we ran multiple linear regression test.

#### NYS County Health Factors and Covid Cases

$N_{Covid\_Cases} \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} Physical\_Environment + \beta_{4} SES\_Factors + \beta_{5} Median\_Age + \beta_{6} Population\_Density + \epsilon$

```{r, message = FALSE, warning = FALSE}
m1 = comb %>% 
  filter(year == '2021') %>% 
  lm(n_cases ~ health_beh_r + clinical_care_r + env_r + ses_r + medage + 
       pop_d, data = .)

m1 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Outbreak"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                full_width = FALSE)
```

After adjusting for median age and population density, only health behaviors ranking was shown to be significantly associated with the number of COVID-19 cases (p < .05). The estimate for health behavior ranking was -2399.531, indicating that counties with higher health behavior ranking had lower number of COVID cases.  

<br>Health behaviors score is based on community residents' tobacco use, diet & exercise, alcohol & drug use, and sexual activity (i.e., number of STI, teen pregnancy). These subfactors are often viewed as drivers of health inequity. Rate of teen pregnancy is higher in majority-minority district; alcohol and drug abuse are more common among minorities; and residential segregation limits minorities access to healthy food. Therefore, the results warrant further exploration of interaction between the health behaviors ranking, its subfactors, and race/ethnicity.

To test the validity of our results, we checked regression assumptions.

```{r m1_check, message = FALSE, warning = FALSE}
par(mfrow = c(2,2))
plot(m1)
```

Test for linearity (residual vs. fitted values) does not show strong pattern; however, the points are not perfectly, randomly scattered around residual = 0. Test for influential observations (residual vs. leverage) shows that there is one influential point (#31), which should be tested to see whether it should be removed or kept. Test for homoscedasticity (scale-location) indicates potential heteroscedasticity. Lastly, test for normality of residuals (normal Q-Q plot) confirms that there is no extreme skewness in our data with  majority of the points falling along the line. Some deviations at the lowest and highest quantiles suggest that random fluctuation may be seen at extreme ends. Combined, the plots indicate that our model, though sufficient, can be improved (i.e., add interaction terms, adjust for more variables).

#### NYS County Health Factors and Covid Fatalities

$N_{Deaths} \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} Physical\_Environment + \beta_{4} SES\_Factors + \beta_{5} Population\_Density + \beta_{6} Median\_Age + \epsilon$
```{r, message = FALSE, warning = FALSE}
m2 = comb %>% 
  filter(year == '2021') %>% 
  lm(n_deaths ~ health_beh_r + clinical_care_r + env_r + ses_r + pop_d + 
       medage, data = .)

m2 %>% 
broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Deaths"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

As seen with model 1, model 2 shows that only health behaviors ranking is significantly associated with the number of COVID deaths in each county (p < .05). The estimate is -43.684, meaning that counties with better health behaviors ranking had less number of COVID deaths. The finding suggests that public health professionals should aim to mitigate disparities in COVID-19 by first starting short-term community-based interventions that encourage residents to build healthier lifestyles and make healthier life choices.

```{r m2_check, message = FALSE, warning = FALSE}
par(mfrow = c(2,2))
plot(m2)
```

Results of regression assumption tests are also similar to that of model 1's, indicating that there is more to be explored in the association between health factor and COVID-19 fatalities. Once again, point 31 shows up as an outlier that needs to be tested further.

#### NYS County Health Factors and Vaccination (Dose 1)

$Vaccine\_Dose1 \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} SES\_Factors + \beta_{4} Physical\_Environment + \beta_{5} Population\_Density + \epsilon$

```{r, message = FALSE, warning = FALSE}
m3 = comb %>% 
  filter(year == '2022') %>% 
  lm(vax_dose1 ~ health_beh_r + clinical_care_r + ses_r + env_r + pop_d, data = .,)

m3 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Vaccination (Dose # 1)"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

In this model, we did not adjust for median age since COVID-19 vaccination rate did not vary significantly for 18+ by 2022. As seen in models 1 and 2, model 3 once again showed that county's health behaviors rank is significantly associated with the number of individuals who received the first dose of the vaccine (p < .05). Once again, the estimate is negative (-11691). This implies that individuals who take care of their health holistically may be more likely to choose to receive the vaccine.

```{r m3_check, message = FALSE, warning = FALSE}
par(mfrow = c(2,2))
plot(m3)
```

Regression assumption tests, once again, show room for improvement in our model and the need to test the outliers, especially point #31.

<p style="color: red">***Question 2: Can race/ethnicity or socioeconomic status in counties help determine COVID-19 severity?***</p>

#### Socioeconomic Factors, Race/Ethnicity and Vaccination Rate

$Vaccination \sim \beta_{0} + \beta_{1} Segregation\_Score + \beta_{2} Median\_Income + \beta_{3} Percent\_Minority + \beta_{4} Population\_Density + \epsilon$


```{r, message = FALSE, warning = FALSE, echo = FALSE}
m4 = comb %>% 
  filter(year == '2022') %>% 
  glm(vax_maj ~ segregation_score + median_income + p_minority + pop_d, data = .)

summary(m4)

m4 %>% 
  broom::tidy() %>% 
  mutate(OR = exp(estimate),
         OR_CI_lower = exp(estimate - 1.96*(std.error)),
         OR_CI_upper = exp(estimate + 1.96*(std.error))) %>% 
  select(term, log_OR = estimate, OR, OR_CI_lower, OR_CI_upper, p.value) %>% 
  kbl(
    caption = "SES Factors & COVID Vaccination"
    , col.names = c("Predictor", "Estimate", "OR", "OR CI, low", "OR CI, high", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```
Based on the the results, median_income and p_minority are significantly associated with higher chance of majority of the county's resident being vaccinated (p < .05). However, despite being statistically significant, because the estimates aare ~0 and ORs are also ~1, the significance is not meaningful. Therefore, though SES factors may contribute, their effects, perhaps, are not as pronounced unless combined with geographical proximity to locations with COVID-19 outbreak. The findings in model 3 could also have happened because COVID vaccine was available at locations such pharmacies (CVS, Walgreen, etc.) and community health centers, removing the barrier to healthcare access and increasing equal access to the vaccine.

Upon checking our model,

```{r m4_check, message = FALSE, warning = FALSE}
par(mfrow = c(2,2))
plot(m4)
```

we did see the need to adjust our model; however, with the lack of 2022 data and demographics information, further analysis and modifications could not be completed. The residual vs. fitted plot did not give us much information as the outcome variable of our model is binary. The normal Q-Q plot showed that our data is not skewed, though there are deviations at extreme quantiles. Furthermore, there appears to be potential heteroscedasticity per scale-location plot and one outlier (point #3 falls outside Cook's distance). 

#### Socioeconomic factors, Race/Ethnicity and COVID Fatalities

$N_{Deaths} \sim \beta_{0} + \beta_{1} Segregation\_Score + \beta_{2} Percent\_Minority + \beta_{3} Median\_Income + \beta_{4} Median\_Age + \beta_{5} Population\_Density + \epsilon$

```{r, message = FALSE, warning = FALSE}
m5 = comb %>% 
  filter(year == '2021') %>% 
  lm(n_deaths ~ segregation_score + p_minority + median_income + medage + pop_d, data = .)

m5 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "SES Factors & COVID Deaths"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

This model shows the number of COVID-19 fatalities in each county is significantly associated with segregation score, % minority (black and Hispanic), and median income (p <= .05) after adjusting for population density. Segregation score had a positive estimate of 17.565, indicating that residential segregation of white vs. non-white could have played a role in exacerbating the COVID-19 disparity. Positive estimate for % minority (estimate = 44.441) reinforces this notion. As for median income, because the estimate is ~0, despite statistical significance, meaningful interepration is lacking. These findings demonstrate that unlike vaccination, racial/ethnic minorities, specifically Hispanic and Black, had a higher risk of dying from COVID compared to White/Asian.

```{r m5_check, message = FALSE, warning = FALSE}
par(mfrow = c(2,2))
plot(m5)
```

For regression test plots, the conclusions are very similar to that of model 4's. The residual vs. fitted plot does not show clear patterns; however, the points are also not equally, randomly distributed along residual = 0 line. The Q-Q plot confirms our data is normally distributed with no extreme skewness. Scale-location shows possibility of homoscedasticity violation and residual vs. leverage, shows that points #3 and #31 are influential observations; combined with model 1 to model 4, this indicates that points 3 and 31 should be identified and tested to determine whether they should be kept or removed from our dataset.

### Model 6
<p style="color: red">***Question 3: Are county percentages of smokers, obesity, physical inactivity, and excessive alcohol drinking associated with receiving at least the first dose of any COVID-19 vaccine?***</p>

The health behaviors ranking was a significant predictor of having increases in the amount of people who have had at least their first dose of the COVID-19 vaccine. The health behaviors ranking encompasses a broad array of sub-variables. Of particular interest to explore are well known factors such as smoking, obesity, exercise, and alcohol consumption. All these factors are well established predictors of many existing health conditions such as CVD, diabetes, cancer, etc. Therefore, we are now interested in seeing if these factor's affect the attainment of getting at least the first dose of the COVID vaccine. 

Note: Again, since the distribution of vaccines for COVID began in 2021 and the complete data for 2022 is not yet available, we will focus our model on data from the 2021 year.

$vax_dose_1 = \beta_{0} + \beta_{1} percent\_smokers + \beta_{2} percent\_obese + \beta_{3} percent\_physically\_inactive + \beta_{4} percent\_excessive\_drinking$

```{r}

subm1 =
  new_comb %>%
  filter(year == '2021') %>%
  lm(vax_dose1 ~ percent_smokers + percent_obese + percent_physically_inactive + percent_excessive_drinking, data = .)

subm1 %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  knitr::kable(digits = 3) 


```


From the results, we can see that every predictor has a significant p-value of less than 0.05, except for the variable `percent_excessive_drinking`. The negative estimates for smokers and obesity suggests that for each percent increase in the adults with obesity or smoking, the amount of people in that county who have at least the first dose of a COVID vaccine decreases. However, for the variables for physical inactivity, the estimate is positive. The estimates in this model are not completely reliable as we have previously established that our COVID outcomes (cases, deaths, and first dose of vaccine) is dependent on geospacial factors. Our data is by county and not individuals which have an influence on the results. Further analysis is necessary on the group level to get more precise estimates. However, from these results, public health professional can focus resources and interventions on smokers and those with obesity to increase vaccination numbers. 


### Model 7

<p style="color: red">***Question 4: Adjusting for median age, is unemployment associated with receiving at least the first dose of any COVID-19 vaccine?***</p>


Unemployment can be a predictor for a variety of outcomes. Often times, people cannot prioritize their health as a result of a lack of money due. It is worth seeing if unemployment is associated with getting the first dose of a COVID vaccine. 


$vax_dose_1 = \beta_{0} + \beta_{1} medage + \beta_{2} percent\_unemployed$

```{r}

subm2 =
  new_comb %>%
  filter(year == '2021') %>%
  lm(vax_dose1 ~ medage + percent_unemployed, data = .)

subm2 %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  knitr::kable(digits = 3) 

```

From the results, we can see that after adjusting for median age, the p-value for unemployment is significant. As stated before, since the outcomes are geospacially correlated, the estimate values are not accurate. But the direction states that for each percent increase in unemployment in a county, the number of people who have had at least one dose of a COVID vaccine decreases. This information can help public health professionals direct resources and attention towards people who are currently unemployed. People who are unemployment may not have gotten the vaccine because they think that it may be an extra expense that they cannot afford to have. This is where it is important to reach out to this group to make sure they are informed that the vaccine is free for everyone and encourage them to get vaccinated.

## Spatial Analysis

```{r}
ny_covid = read.csv('data/covid_testing.csv')  %>%
  janitor::clean_names()

###############################
ny_covid$test_date = lubridate::mdy(ny_covid$test_date)
##############################
total <- merge(ny_covid,NY_df,by="county")


```


```{r}

mon = seq(as.Date("2021-01-01"), as.Date("2022-01-01"), by = "3 months")
df_list = list()
# loop version 2
for (i in 1:4) {
  sub_df = total[total$test_date >= mon[[i]] & total$test_date <= mon[[i+1]], ]
  group_df =  sub_df %>% group_by(county)  %>%
    summarise(total_covid = sum(new_positives))
  df = dplyr :: left_join(group_df, NY_df, by = 'county')
  df_list[[i]] = df
}


p <- list()
for(i in 1:4){
  p[[i]] <- df_list[[i]] %>%
    ggplot() +geom_sf(aes(fill = total_covid, geometry = geometry)) + 
    scale_fill_viridis_c(option = "viridis") 
}

do.call(grid.arrange,p)

```
The above 4 chloroplast maps are a summary 

## Autoregression

```{r}
ny_covid = read.csv('data/covid_testing.csv')  %>%
  janitor::clean_names()

######################################################
ny_covid$test_date = lubridate::mdy(ny_covid$test_date)
sub_df = ny_covid[ny_covid$county == 'New York City', ]
sub_df<- sub_df[seq(dim(sub_df)[1],1),]

ggplot(sub_df, aes(x=test_date, y=new_positives)) + 
  geom_point(alpha = 10/100, colour = "red") + 
  scale_x_date(date_labels = "%m-%Y") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "New positive  covid cases per month in NYC")

```
The above plot is a scatter plot of the daily positive cases in New York City from March 2020 up-till now. We can see the wave like trend, especially from hill shaped curves at the start of the 2020, start of 2021, start of 2022 when it peaked around mid 2022. This implies the existence of auto regressive behavior which we were interested to investigate. To better understand the auto-regressive trend we built a Autocorrelation Function (ACF) plot and Partial Autocorrelation Function (PACF) plot. The ACF and PACF plots are used to figure out the order of AR, MA, and ARMA models. Autocorrelation is the correlation between a time series with a lagged version.
The partial autocorrelation at lag k is the autocorrelation between X_t_t and X_(t-k) that is not accounted for by lags 1 through ð‘˜âˆ’1. 


```{r}
library(forecast)
library(ggplot2)

conf.level <- 0.95
ciline <- qnorm((1 - conf.level)/2)/sqrt(length(sub_df$new_positives))

############# ACF ####################
bacf <- acf(sub_df$new_positives, plot = FALSE)
bacfdf <- with(bacf, data.frame(lag, acf))

p1 <- ggplot(data=bacfdf, mapping=aes(x=lag, y=acf)) +
       geom_bar(stat = "identity", position = "identity") +
  ggtitle("Autocorrelation Plot")


########### PACF ###################
bacf <- pacf(sub_df$new_positives, plot = FALSE)
bacfdf <- with(bacf, data.frame(lag, acf))


p2 <- ggplot(data=bacfdf, mapping=aes(x=lag, y=acf)) +
  geom_bar(stat = "identity", position = "identity",fill="lightblue")  +
  ggtitle("Partial Autocorrelation Plot")

grid.arrange(p1,p2)

```
Both the ACF and PACF start with a lag of 0, which is the correlation of the time series with itself and therefore results in a correlation of 1. However after a lag of 1, ACF gives us values of auto-correlation of any series with its lagged values whereas PACF finds correlation of the residuals that remains after removing the effects which are already explained by the earlier lag(s). Thus PACF might be a better indicator of how many lag terms to keep. In our case we decided to move with the 8 lag terms since they seem to have significant partial autcorrelation with value at lag = 0 (present value).

```{r}
reg1 = arima(sub_df$new_positives, order = c(8,0,0))
summary(reg1)
```
The coefficients for an AR(8) model represent the weights that are applied to the past 8 values of the time series (i.e. $y_{t-1}$, $y_{t-2}$, $y_{t-3}$, $y_{t-4}$, $y_{t-5}$, $y_{t-6}$, $y_{t-7}$, $y_{t-8}$) in order to predict the next value of the time series ($\hat{y}_t$). In this case, the coefficients are as follows:

These coefficients can be used to write the ARIMA (8,0,0) equation as follows:

$\hat{y}t = \alpha_1 y{t-1} + \alpha_2 y_{t-2} + \alpha_3 y_{t-3} + \alpha_4 y_{t-4} + \alpha_5 y_{t-5} + \alpha_6 y_{t-6} + \alpha_7 y_{t-7} + \alpha_8 y_{t-8} + \beta_0$

$\begin{aligned} \hat{y}t &= 1.0562 \cdot y_{t-1} - 0.2312 \cdot y_{t-2} + 0.0948 \cdot y_{t-3} + 0.0779 \cdot y_{t-4} - 0.1035 \cdot y_{t-5} + 0.1343 \cdot y_{t-6} + 0.3918 \cdot y_{t-7} - 0.4517 \cdot y_{t-8} + 2924.354 \end{aligned}$

This equation can be used to predict the next value of the time series, $\hat{y}_t$, based on the past 8 values of the time series. The coefficients in the equation represent the relative importance of each of the past values in predicting the next value.

Plotting the fitted values alongside the positive covid cases:
```{r}
ar = fitted(reg1)
ggplot(sub_df, aes(x=test_date, y=new_positives)) + 
  geom_point(alpha = 20/100, color = "red") + 
  scale_x_date(date_labels = "%m-%Y",date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "Scatter Plot of Daily Postive Cases in New York City from 2020-2022
       with ARIMA trend line", x= "Test Date",
       y = 'Positive Cases Detected') +
  geom_line(aes(y=ar,color = "ARIMA trend line")) +  
  scale_color_manual(values = c("ARIMA trend line" = "black"),
                                          labels = "Fitted Arima Model") 

```


## Discussions 





## Limitations 




## Conclusion 