---
title: "Health Factors EDA"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 4
    theme: lumen
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
library(tidyverse)
library(tidytext)  
library(httr)
library(rvest)
library(lubridate)
library(plotly)
library(patchwork)
library(lmtest)
library(dplyr)
library(tidycensus)
require(data.table)
library(gridExtra)
library(ggplot2)
library(data.table)
library(tidycensus)
library(tidyverse)
library(ggridges)
library(patchwork)
library(viridis)
library(modelr)
library(mgcv)
library(readxl)
library(plotly)
library(kableExtra)
library(Hmisc)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "100%")

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r census_data, results = FALSE, message = FALSE}
census_api_key("aa512886c5449a582d837da8d3a07af66a043fe5")

census_data <- load_variables(2020, "acs5", cache=T)
fwrite(census_data, "census_variables.csv")

vars <- c(tpop = 'P001001',
          medage = 'P013001',
          wpop = 'P003002',
          bpop = 'P003003',
          apop = 'P003005',
          hpop = 'P004003')

nys_df <- get_decennial(state = "ny", 
                       geography = "county",
                       variables = vars,
                       geometry = T,
                       output = "wide")

NY_df <- get_decennial(state = "ny", 
                       geography = "county",
                       variables = vars,
                       geometry = T,
                       output = "wide")

NY_df = NY_df %>%
  mutate(
    county = gsub(" County, New York","",NAME))

nys_df = nys_df %>%
  mutate(
    county = gsub(" County, New York","",NAME))

nys_df$race_div <- 1 - (((nys_df$wpop*(nys_df$wpop-1))+
                          (nys_df$bpop*(nys_df$bpop-1))+
                          (nys_df$hpop*(nys_df$hpop-1))+
                          (nys_df$apop*(nys_df$apop-1)))/
                         (nys_df$tpop*(nys_df$tpop-1)))

nys_df = nys_df %>% 
  select(tpop, medage, county, race_div)


area = read_excel('data/pop_density.xlsx') %>% 
  janitor::clean_names() %>% 
  rename(area = value)

nys_df = left_join(nys_df, area, by = 'county') %>% 
  mutate(pop_d = tpop/area)


# 2019 HR Data

hr2019 = read_excel('data/ny_hr19.xls', sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2019 = read_excel('data/ny_hr19.xls', sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'segregation_index_2',
         'household_income',
         'percent_hispanic',
         'percent_african_american') %>% 
  rename(median_income = household_income,
         segregation_score = segregation_index_2,
         p_hispanic = percent_hispanic,
         p_black = percent_african_american) %>%
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var19 =
  read_excel('data/ny_hr19.xls', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_obese, food_environment_index, percent_physically_inactive, percent_with_access, percent_excessive_drinking, percent_alcohol_impaired, number_chlamydia_cases, teen_birth_rate, percent_unemployed)
  

list_2019 =
  list(hr2019, demo2019, sub_var19)

total2019 = 
  list_2019 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2019)

# 2020 HR Data
hr2020 = read_excel('data/ny_hr20.xlsx', sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2020 = read_excel('data/ny_hr20.xlsx', sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'median_household_income', 
         'segregation_index_2',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index_2,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var20 =
  read_excel('data/ny_hr20.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)
  

list_2020 =
  list(hr2020, demo2020, sub_var20)

total2020 = 
  list_2020 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2020)

# 2021 HR Data
hr2021 = read_excel('data/ny_hr21.xlsx', sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2021 = read_excel('data/ny_hr21.xlsx', sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select('county', 
         'median_household_income', 
         'segregation_index_2',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index_2,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_black + p_hispanic)

sub_var21 =
  read_excel('data/ny_hr21.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)

list_2021 =
  list(hr2021, demo2021, sub_var21)

total2021 = 
  list_2021 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2021)

# 2022 HR Data
hr2022 = read_excel('data/ny_hr22.xlsx', 
                    sheet = 'Outcomes & Factors SubRankings', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
  select(-starts_with("z"), -fips, -state) %>% 
  rename(longevity_r = rank_5,
         qol_r = rank_7,
         health_beh_r = rank_9,
         clinical_care_r = rank_11,
         ses_r = rank_13,
         env_r = rank_15)

demo2022 = read_excel('data/ny_hr22.xlsx', sheet = 'Additional Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>% 
    select('county', 
         'median_household_income', 
         'segregation_index',
         'percent_black',
         'percent_hispanic') %>% 
  rename(median_income = median_household_income, 
         segregation_score = segregation_index,
         p_black = percent_black,
         p_hispanic = percent_hispanic) %>% 
    mutate(segregation = case_when(segregation_score < 30 ~ 0,
                                 segregation_score >= 30 ~ 1),
           p_minority = p_hispanic + p_black)

sub_var22 =
  read_excel('data/ny_hr22.xlsx', sheet = 'Ranked Measure Data', skip = 1) %>% 
  janitor::clean_names() %>% 
  slice(-c(1)) %>%
  select(county, percent_smokers, percent_adults_with_obesity, food_environment_index, percent_physically_inactive, percent_with_access_to_exercise_opportunities, percent_excessive_drinking, percent_driving_deaths_with_alcohol_involvement, number_chlamydia_cases, teen_birth_rate, percent_unemployed) %>%
  rename(percent_obese = percent_adults_with_obesity,
         percent_with_access = percent_with_access_to_exercise_opportunities,
         percent_alcohol_impaired = percent_driving_deaths_with_alcohol_involvement)

list_2022 =
  list(hr2022, demo2022, sub_var22)

total2022 = 
  list_2022 %>%
  reduce(full_join, by = "county") %>%
  mutate(year = 2022)

# Merge
hr_all = rbind(total2019, total2020, total2021, total2022)


ny_test = read_csv('data/ny_covidtest.csv') %>% 
janitor::clean_names() %>% 
filter(geography == 'COUNTY') %>% 
select(-geography) %>% 
rename(date = test_date) %>% 
mutate(date = lubridate::mdy(date))  

ny_death = read_csv("data/ny_coviddeaths.csv") %>% 
janitor::clean_names() %>% 
rename(date = report_date, n_fatality = place_of_fatality) %>%
select(-deaths_by_county_of_residence) %>% 
mutate(date = lubridate::mdy(date), county = replace(county, county == 'Manhattan', 'New York'))

ny_vax = read_csv("data/ny_covidvax.csv") %>% 
janitor::clean_names() %>% 
rename(date = report_as_of) %>% 
select(-region, -series_complete) %>% 
mutate(date = lubridate::mdy(date))

ny_covid_v1 = left_join(ny_test, ny_death, by = c("date", "county"))

covid_df_v1 = left_join(ny_covid_v1, ny_vax, by = c("date", "county")) %>% 
  separate(date, c("year", "month", "day"), sep = "-") %>% 
  mutate(year = as.numeric(year), month = as.numeric(month), day = as.numeric(day))

covid_df = covid_df_v1 %>% 
  group_by(year, county) %>% 
  summarise(n_cases = sum(new_positives),
            vax_dose1 = max(first_dose),
            n_deaths = max(n_fatality, na.rm = TRUE))


covid_hr = left_join(covid_df, hr_all, by = c("county", "year")) 
comb = left_join(covid_hr, nys_df, by = 'county') %>% 
  mutate(tpop_50 = tpop/2,
         vax_maj = case_when(vax_dose1 >= .83*(tpop) ~ 1,
                             vax_dose1 < .83*(tpop) ~ 0))
```

For our exploratory data analysis, we wanted to see if there is a pattern to being of a certain rank in each of the four rankings (health behaviors, clinical care, SES, and physical environment) and the number of COVID cases, deaths, and vaccinations. For each health factor ranking, we made quartiles to group the rankings, with Quartile 1 having the best rankings and Quartile 4 having the worst rankings.

To graph each plot, we had to create a new dataframe, from the `comb` final dataset, and used `pivot_longer` to separate the ranking variables into 2 different variables: the rank name and the rank number; we also renamed the ranking variable. We also created a new `quartile` variable for each ranking that assigns each observation to one of four quartiles based on ranks. 

Again, we filtered out only 2021 values for each county as there is only complete vaccination information available for the year 2021.   

```{r}
eda1 = 
  comb %>% 
  rename("Clinical Care" = clinical_care_r,
         "Physical Environment" = env_r,
         "Health Behaviors" = health_beh_r,
         "Social and Economic Factors" = ses_r) %>% 
  pivot_longer("Health Behaviors":"Physical Environment",
               names_to = "rank_var",
               values_to = "rank") %>% 
  filter(year == '2021') %>% 
  mutate(Quartile = case_when(rank < 17 ~ 'Quartile 4',
                              rank < 32 ~ 'Quartile 3',
                              rank < 46 ~ 'Quartile 2',
                              rank < 63 ~ 'Quartile 1')) %>% 
  mutate(Quartile = as.factor(Quartile)) 
```

## COVID Deaths, Cases, and Vaccination

```{r}
p1 = ggplot(data=eda1, aes(x=rank_var, y=n_deaths, fill=Quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Reds") + 
  theme_minimal() +
  labs(title="Number of COVID Deaths and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of Deaths")

p2 = ggplot(data=eda1, aes(x=rank_var, y=n_cases, fill=Quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Oranges") + 
  theme_minimal() +
  labs(title="Number of COVID Cases and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of Cases")

p3 = ggplot(data=eda1, aes(x=rank_var, y=vax_dose1, fill=Quartile)) +
  geom_bar(stat="identity", position=position_dodge())+
  scale_fill_brewer(palette="Greens") + 
  theme_minimal() +
  labs(title="Covid Vaccination (Dose 1) and County Health Factors by Quartile",
        x ="County Health Factors", y = "Number of People Vaccinated")

grid.arrange(p1, p2, p3, ncol = 1)
```

From the results of the plots, we can see that there is a general pattern across all the COVID-19 outcomes (cases, deaths, and first dose of a vaccine) for the Health Behaviors Rankings. Counties with higher (worse) ranks show a higher number of cases, deaths, and first dose of vaccines. We will explore the variables that contribute to this pattern in our statistical analysis.


