---
title: "County Health Factors Ranking & COVID-19 Analysis"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    toc_depth: 4
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(kableExtra)

source("code/data_cleaning.R")
```

## County Health Factors and COVID-19

In the data visualization for HF and COVID-19, we observe patterns across all three COVID-19 variables: fatalities, cases, and vaccination for only health behaviors. Therefore, we decided to explore this further and test whether this holds true even after adjusting for population density and median age, which were often shown to be strongly correlated and/or associated with COVID-19.

### COVID Cases

First, we tested for association among HF (health behavior, clinical care, physical environment, and socioeconomic factors) and COVID-19 cases after adjusting for median age and population density in each county. 

$N_{Covid\_Cases} \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} Physical\_Environment + \beta_{4} SES\_Factors + \beta_{5} Median\_Age + \beta_{6} Population\_Density + \epsilon$

```{r, message = FALSE, warning = FALSE}
m1 = comb %>% 
  filter(year == '2021') %>% 
  lm(n_cases ~ health_beh_r + clinical_care_r + env_r + ses_r + medage + 
       pop_d, data = .)

m1 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Outbreak"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

Among the 4 factors, only health behaviors ranking's p-value is < .05, indicating that health behaviors ranking is significantly associated with the number of COVID cases.

### COVID Deaths

$N_{Deaths} \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} Physical\_Environment + \beta_{4} SES\_Factors + \beta_{5} Population\_Density + \beta_{6} Median\_Age + \epsilon$

```{r}
m2 = comb %>% 
  filter(year == '2021') %>% 
  lm(n_deaths ~ health_beh_r + clinical_care_r + env_r + ses_r + 
       pop_d + medage, data = .)

m2 %>% 
broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Deaths"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

For COVID fatalities, once again, only the p-value for health behaviors ranking is statistically significant (p < .05).

### COVID Vaccination

$Vaccine\_Dose1 \sim \beta_{0} + \beta_{1} Health\_Behavior + \beta_{2} Clinical\_Care + \beta_{3} SES\_Factors + \beta_{4} Physical\_Environment + \beta_{5} Population\_Density + \epsilon$

```{r}
m3 = comb %>% 
  filter(year == '2022') %>% 
  lm(vax_dose1 ~ health_beh_r + clinical_care_r + ses_r + env_r 
     + pop_d, data = .,)

m3 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Factor Ranking & COVID Vaccination (Dose # 1)"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

The results indicate that only health behaviors ranking is significantly associated with the number of individuals who received the first dose of the vaccine by 2022 (p < .05).

### Subfactor Analyses

**Health Behavior Subfactors & Covid Vaccination (Dose 1)**

$vax_dose_1 = \beta_{0} + \beta_{1} percent\_smokers + \beta_{2} percent\_obese + \beta_{3} percent\_physically\_inactive + \beta_{4} percent\_excessive\_drinking$

```{r}
subm1 =
  new_comb %>%
  filter(year == '2021') %>%
  lm(vax_dose1 ~ percent_smokers + percent_obese + percent_physically_inactive 
     + percent_excessive_drinking, data = .)

subm1 %>% 
  broom::tidy() %>%
  select(term, estimate, p.value) %>% 
  kbl(
    caption = "Health Behavior Subfactors and COVID Vaccination"
    , col.names = c("Predictor", "Estimate", "p-value")
    , digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

The regression test showed that % smokers, obese, and physically inactive are significantly associated with the number of individuals who have received the first dose of COVID vaccine by 2022 (p < .05). Meanwhile, % excessiver drinkers was not statistically significant (p = .41).